---
title: "The Impact of Hurricane Maria on Out-Migration From Puerto Rico"
author: "Aryaa Rajouria, Crystal Yu"
date: "May 31 2022"
output: 
  bookdown::html_document2:
        number_sections: true
        theme: flatly
        self_contained: true
        toc: true
        toc_float:
            collapsed: false
            smooth_scroll: false
        code_folding: hide
  editor_options: 
          markdown: 
          wrap: 72


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

```

# **Introduction & Set-up** 

Today, we will be going through [The Impact of Hurricane Maria on Out-migration from Puerto Rico: Evidence from Facebook Data](https://www.jstor.org/stable/pdf/45216967.pdf). The GitHub Repository can be found [here](https://canvas.uw.edu/courses/1549181/modules/items/15393060). The GitHub Repository for this tutorial can be found [here](https://github.com/aryaa-rajouria/FB_Migration_Hurricane_Maria_Tutorial).

In this paper, Alexander et al use Facebook advertising data to estimate migration of Puerto Ricans to the continental United States after Hurricane Maria. Given the large scale use of Facebook in Puerto Rico, they argue that using Facebook data is appropriate in this instance. They collect Facebook demographic data every 2-3 months (they have six waves in total) from January 2017 to March 2018. Hurricane Maria took place between September 2017 and October 2017, enabling the authors to see the effect of the hurricane on the population demographics of Puerto Rico. They use a difference-in-differences model to compare the size of the Puerto Rican and continental US populations before and after the hurricane, finding that between October 2017 and January 2018, there was a 17% increase in the number of migrants from Puerto Rico the the continental United States--with the most migrating to Florida. They also find that there were more male migrants than females, with many being younger, i.e. in working age groups. They also find that after the hurricane, 1.8% of migrants returned to Puerto Rico. The findings aligned with prior research. 

A few examples of instances in which the Facebook Ads Manager has previously been used include: 

1. [Tracking cross-border migration in Europe during the Coronavirus pandemic](https://globalizationandhealth.biomedcentral.com/articles/10.1186/s12992-022-00832-6)
2. [Using FacebookAds for election polling](https://dl.acm.org/doi/abs/10.1145/3323503.3349552)
3. [To create estimates to improve global development statistics](https://www.itu.int/dms_pub/itu-s/opb/journal/S-JOURNAL-ICTF.VOL1-2018-2-P04-PDF-E.pdf)
4. [To create global COVID-19 Surveys to aid in policy-making](https://csde.washington.edu/seminar/partnering-with-a-global-platform-toinform-research-and-public-policy-making-what-needs-to-be-in-place-to-make-a-global-covid-19-survey-work/) which was presented at the CSDE seminar series last year; the presentation can be viewed on [YouTube here](https://www.youtube.com/watch?v=t_hvQEYK9Xg).

In this R Tutorial, we will be using the data Alexander et al. provide on their GitHub for replicability. We will not be extracting new data from the Facebook API. However, if you are interested in learning how to do that, this is the general process: 

1. Access the Web Interface of the Marketing API [here](https://www.facebook.com/adsmanager/manage/campaigns?act=147920879) and define relevant attributes.
2. Create a developer account and obtain an [access token](https://developers.facebook.com/tools/accesstoken) as a Facebook developer.
3. Access Facebook data that is publicly [available](https://developers.facebook.com/docs/graph-api).

For more information on how to extract data from the Facebook Ads Manager API using Python see [this guide](https://worldbank.github.io/connectivity_mapping/facebook_nbs/pipeline.html) and the information [here](https://developers.facebook.com/docs/graph-api/overview/), [here](https://developers.facebook.com/blog/post/2017/04/18/graph-api-v2.9/), and [here](https://towardsdatascience.com/how-to-use-facebook-graph-api-and-extract-data-using-python-1839e19d6999). The [Facebook Developers](https://developers.facebook.com) website provides comprehensive directions. When using access the Web Interface of the Marketing API using Python, you will need to install pySocialWatcher, which can be found on [GitHub](https://github.com/maraujo/pySocialWatcher).

## Load in Necessary Libraries 

First, we will download the necessary libraries.

```{r, message=FALSE, warning=FALSE}
## load libaries
library(kableExtra)      # for making pretty tables
library(RColorBrewer)    # for color palettes
library(lubridate)       # for time data
library(rnaturalearth)   # for world shapefile
library(sf)              # for simple feature mapping
library(mapview)         # for simple interactive mapping
library(tmap)            # for interactive mapping
library(tidyverse)       # for data wrangling  
library(viridis)         # for plot colors
library(migest)          # for migration visualization 
library(dplyr)           # because Aryaa's quirky R needs her to explicitly call it for it read in pipes 

```

# **Facebook Data** 

## Loading in the Facebook Data

Next, we will load in the FB data from GitHub, wrangle the data into the format we want, and add in wave information. 

```{r, results='hide'}
## load the raw facebook data
#fb.raw <- read_csv("./data/facebook/master_acs_facebook_data.csv")  # if you've downloaded or cloned the github repo, otherwise
fb.raw <- read_csv("https://raw.githubusercontent.com/MJAlexander/fb-migration-hurricane-maria/master/data/facebook/master_acs_facebook_data.csv") 

dim(fb.raw)  # 25,200 rows, 30 columns

## clean & formatting the data
fb <- fb.raw %>%
  # extract age group from character string (ex ages_15_19)
  mutate(age_group = as.numeric(unlist(lapply(strsplit(age_group, "_"), '[[', 2)) ) ) %>%
  #selecting relevant variables 
  select(state, origin, age_group, sex, expat_population_wave1:facebook_population_wave7) %>%
  # add wave indicator
  pivot_longer(5:18, names_to="var", values_to="value") %>% 
  mutate(wave = as.numeric(substr(var, str_length(var), str_length(var)) ),
         var = substr(var, 1, str_length(var)-6)) %>%
  pivot_wider(names_from=var, values_from=value) %>%
  # "fix" state names (states with a space in the name, e.g. new york, are coded with an underscore in the name, e.g. new_york)
  mutate(state = ifelse(grepl("_", state), str_replace(state, "_", " "), state))
  
## add in corresponding dates for each wave: 
##   requires lubridate package (part of the tidyverse, for date/time data, but need to separately load library)
#require(lubridate)
fb <- fb %>% 
  mutate(date = case_when(
    wave==1~ymd(20170101),
    wave==2~ymd(20170401),
    wave==3~ymd(20170601),
    wave==4~ymd(20171001),
    wave==5~ymd(20180101),
    wave==6~ymd(20180301),
    wave==7~ymd(20180701)
  ))

```

```{r}
## look at cleaned fb data
#dim(fb)  # 176,400 rows, 8 columns: 7 waves x 2 sex x 9 age groups x 50 states x 28 origin countries
   head(fb)# note the date format on the date column; not numerical but in data format 
```

The FB data is in long format, and has an observation (row) for each wave-state-country of origin-sex-age group combination. There are a total of 176,400 rows: one for each combination of wave (7), sex (2), 5-year age groups (9), states (50; DC is not included), and country of origin (28).

Data note: The data from waves 6 and 7 contain a significant number of "1000"s in the `expat_population` column (estimated number of migrants), which is suggestive of bottom-coding. In earlier waves, it appears the corresponding bottom-code count was "20". It is possible the large numbers of "20"s in the earlier waves, and "1000"s in later waves, may be hiding other potential data issues. In general, the FB `expat_population` and `facebook_population` estimates clearly involve some rounding.

We will now share some descriptives and visualizations to better understand the clean data. Each new 'wave' of data is collected on the first of the month, every couple of months.

Because the data contains an observation for each state-country of origin-sex-age group combination, and there is evidence of bottom coding and data suppression, the estimates below are biased. This is especially obvious from the estimates for waves 6 and 7.

## Total US population

Alexander et al use FB data to estimate the size of the US population. From their estimations, there seems to be a general decrease in the coverage of the US population from Wave 1 to Wave 7, with an particular decrease in Wave 5. These estimations are clearly an undercount compared to population estimates by the US Census Bureau. 

In 2017, the US population was around 325 million according to the US Census Bureau [American Community Survey](https://data.census.gov/cedsci/table?tid=ACSDP1Y2017.DP05) and [Population Estimates Program](https://data.census.gov/cedsci/table?q=PEPANNRES%3A%20ANNUAL%20ESTIMATES%20OF%20THE%20RESIDENT%20POPULATION%3A%20APRIL%201,%202010%20TO%20JULY%201,%202019%20-%20FOR%20FULL%20ESTIMATES%20DETAIL,%20VISIT%20https%3A%2F%2Fwww.census.gov%2Fprograms-surveys%2Fpopest.html&tid=PEPPOP2019.PEPANNRES).


```{r}
## estimate total facebook population by wave
# filtering by Mexico but it doesn't matter - all countries are the same total population
# (i.e. the facebook_population count is unique only for each specific combination of age-sex-state-wave)

# this is the total (fb) population per state-age-sex-wave combination
total_fb_pop <- fb %>% filter(origin=="Mexico") %>% group_by(wave, date) %>% summarise(fb = sum(facebook_population))

#creating a table of the total fb pop
kable(total_fb_pop, 
      format.args=list(big.mark=','), col.names=c("wave", "date", "estimated US pop")) %>%
  kable_paper()

#using ggplot to visualize fb pop estimations by wave
ggplot(data=total_fb_pop, aes(x=wave, y=fb/1e6)) +
  geom_line()+
  geom_point()+
  labs(title="Estimated US Population Using FB data",x="Waves", y = "Population Estimate in millions") +
  scale_x_continuous(name = "wave", breaks=seq(1:7)) +
  theme(panel.grid.minor.x = element_blank())

```

```{r, eval=FALSE}
# alternatively the estimated fb population can be calculated using the following
fb %>%
  # remove duplicate entries
  distinct(state, age_group, sex, wave, facebook_population, date) %>%
  # calculate estimated population by wave
  group_by(wave) %>%
  summarize(facebook_population = sum(facebook_population), .groups='drop')

```


## Total expats
Similarly, the Facebook data can be used to estimate the size of the immigrant population in the US (Facebook calls this the expat population).  As is evident from the table below, the number of migrants in the US is steady from Wave 1 to Wave 5. There is a decrease from Wave 4 to Wave 5. Wave 6 and Wave 7 are more similar to each other. 

```{r}
#looking at migrant count by date: helps us see what 'dates' are what 'waves'
mig_count <- fb %>% group_by(wave, date) %>% summarise(n = sum(expat_population))
kable(mig_count, 
      format.args=list(big.mark=','), col=c("Wave", "Date", "Count")) %>%
  kable_paper()
```

## Expats by age

Now, we will examine the age patterns of migrants within the different waves. The Facebook data only contains information on individuals between the ages of 15 to 59. As per the table and graph below, of all the migrants in the Facebook data, there were more migrants in their mid-20s and mid-30s compared to younger (teenagers) or older (past 50) populations. All of the waves had similar age patterns. 

```{r}
#looking at how many people who migrated were of what age group generally 
age_comparison <- fb %>% group_by(age_group, wave, date) %>% summarise(n = sum(expat_population))

kable(age_comparison %>%
        mutate(wave = paste("wave", wave, date, sep="_")) %>%
        select(-date) %>%
        pivot_wider(names_from=wave, values_from=n), 
      format.args=list(big.mark=','), 
      caption="estimated age distribution of the migrant population in the US, based on FB advertising data") %>%
  kable_paper()

ggplot(age_comparison, aes(x=age_group, y=n, color=factor(wave)))+
  geom_line() +
  #geom_bar(position="dodge", stat= "identity", fill="#69b3a2") +
  labs(x = "Age", y = "Count", color="wave")

```


## Expats by sex

When we disaggregate the data by different waves, we find that there are differences in the migration experiences of males and females. In general, there are more male migrants than female migrants. These differences are slightly accentuated in certain waves (Wave 3, Wave 5 and Wave 6) but are generally similar in all waves. 

To do this, we group by wave and sex, adding the number of migrants (expat_population) in the FB data set. We then use ggplot2 to create a bar graph to visualize these differences. 

```{r}
#looking at how many people who migrated were male vs female according to wave 
sex_wave_comparison <- fb %>% filter(origin=="Puerto Rico") %>% group_by(wave, sex) %>% summarise(n = sum(expat_population))

#graphing waves 
sex_wave_comparison$wave_factor<- factor(sex_wave_comparison$wave)
sex_wave_comparison$sex_character<- as.character(sex_wave_comparison$sex)
ggplot(sex_wave_comparison, aes(fill=sex_character, y=n, x=wave)) + 
  geom_bar(position="dodge", stat="identity") +
  scale_fill_brewer(palette="Dark2") +
  labs( x = "wave", y = "count", fill = "sex") +
  coord_flip()+
  scale_x_discrete(name = "wave", limits=c(1,2,3,4,5,6,7)) +
  scale_fill_discrete(name = "sex", labels = c("male", "female"))

```

## Expats by country of origin {#expatpopsize}

The Facebook data contains estimates of expats from 28 countries. Of course, this is not representative of all of the countries migrants come from. 

```{r}
## unique expat groups (country of origin)
(fb.origins <- unique(fb$origin))  # there are only 28, as noted in the paper

```

Broken down by country of origin, we see the largest number of migrants in the US are of Mexican origins. This is true for all 7 waves (from January 2017 to July 2018). Notably, the number of Mexican migrants is significantly larger than the estimated migrant population from the other 27 origin countries. 

There is a marked a marked difference in the number of migrants from Puerto Rico between Waves 3 and 4, 4 and 5, and beyond Wave 5.


```{r}
## size of expat groups in the us, according to the fb data
fb.expat <- fb %>% #filter(wave <= 5) %>%  # the counts for waves 6 and 7 look very different when compared to wave 5
  group_by(origin, wave) %>% 
  summarise(expat_pop = sum(expat_population), .groups='drop') %>% 
  arrange(wave, -expat_pop) %>%
  mutate(wave = paste0("pop_wave", wave)) %>%
  pivot_wider(names_from=wave, values_from=expat_pop) 

fb.expat %>%
  kable(format.args=list(big.mark=','), caption="estimated migrant population in the US by country of origin, based on FB advertising data") %>%
  kable_paper("hover")

```

Data note: There is a large difference in counts between Waves 5 and 6 (and similarity between Waves 6 and 7). This illustrates that Facebook data isn't always stable: it can suddenly change. Alexander et al note that a limitation is the opaqueness of the Facebook algorithm or shift in privacy policies. For example, it is interesting to note that the Cambridge Analytica data scandal (where the personal data belonging to millions of Facebook users was collected without their consent) occurred in March 2018 (during Wave 6). Perhaps the increase in media attention regarding social media privacy policies spurred Facebook to change their algorithms, shift their estimates, or change how data is shared in the Ads manager (i.e. bottom-coding, rounding).

We now create a chord map to visualize the migration patterns of those in the FB data set. Examining just the 3 largest migrant groups in the US, as well as Puerto Rico, for wave 4 (October 2017), we see that 

* California and Texas are especially popular destinations for those from **Mexico**
* California is a popular relocation site among those from **India**
* California is the top destination among immigrants from the **Philippines**
* Florida is the top destination for immigrants from **Puerto Rico**

```{r}
#selecting relevant variables: you need these three for a chord chart
# this is wave 4 only
mig_flow <- fb %>% filter(wave == 4 & origin %in% c("Mexico", "India", "Philippines", "Puerto Rico")) %>% group_by(state, origin) %>% summarize(expat_population = sum(expat_population), .groups='drop') 

#putting our data into the mig_chord function 
mig_flow %>% mig_chord()

```

Focusing on just migrants from Puerto Rico, we see that in wave 4 (Oct 2017), California, Connecticut, Florida, Georgia, Illinois, Massachusetts, New Jersey, New York, North Carolina, Ohio, Pennsylvania, Texas, Virginia, and Wisconsin appear to be popular destinations. 

```{r}
# wave 4 pr
mig_flow4 <- fb %>% filter(wave==4, origin== "Puerto Rico" ) %>% #, state=="Texas"|state=="Florida"|state=="New Jersey"|state=="New York"|state=="Pennsylvania"|state=="Connecticut"|state=="Illinois"|state=="California"|state=="Massachusetts"|state=="Washington") %>% 
  select(origin, state, expat_population)

#putting our data into the mig_chord function 
mig_flow4 %>% mig_chord()

```

When we compare the migration patters to Wave 5 (January 2018), we see similar patterns. Hurriciane Maria occurs between Waves 4 and 5. While not evident in the chord maps, when we look at the numbers, we find a slight increase in Puerto Rican migrants to Florida between Wave 4 and Wave 5. In the same time frame, we see a slight decrease in migrants to California and Texas.  

```{r}
## wave 5
mig_flow5 <- fb %>% filter(wave==5, origin== "Puerto Rico") %>%#, state=="Texas"|state=="Florida"|state=="New Jersey"|state=="New York"|state=="Pennsylvania"|state=="Connecticut"|state=="Illinois"|state=="California"|state=="Massachusetts"|state=="Washington") %>% 
  select(origin, state, expat_population)

mig_flow5 %>% mig_chord()

```

## WA FB total population + expat population

Here, we use the FB data to see what it says about the migrant population in Washington. First, we estimate total migrant population. Then, we take a closer look at the migrant population in Washington by country of origin.  

When estimating migrant population in Washington, we find a steady increase from Wave 1 to Wave 5, while the total population in WA is declining. Then, there is a sharp increase in the migrant population between Wave 5 and Wave 6. This suggests that the total count data may not be very reliable for smaller migrant groups, illustrating a potentially significant barrier in using FB data for research. As a result, we leave out waves 6 and 7 when estimating the migration population in Washington according to country of origin. While there are slight fluctuations over time, a majority of the migrants are coming from Mexico, India, Vietnam, China, and Canada. The number of migrants coming from Mexico to Washington are significantly larger than those from other countries.  

```{r}
## estimated wa population by wave, according to fb
wa.fb <- fb %>%
  filter(state == "Washington") %>%
  # remove duplicate entries
  distinct(state, age_group, sex, wave, facebook_population, date) %>%
  group_by(wave, date) %>%
  summarize(facebook_population = sum(facebook_population), .groups='drop')

## estimated expat population in wa, according to fb
  # notice there's a big jump in numbers between waves 5 and 6
fb %>% 
  filter(state == "Washington") %>%
  group_by(wave) %>%
  summarize(expat_pop = sum(expat_population)) %>%
  # merge in fb population for wa
  left_join(wa.fb, by="wave") %>%
  mutate(expat.prop = round(expat_pop/facebook_population, 3)) %>%
  relocate(date, .after=wave) %>%
  kable(format.args=list(big.mark=','), caption="estimated population sizes in WA, based on FB advertising data",
        col.names=c("wave", "date", "estimated WA expat pop", "estimated WA total pop", "proportion WA expat")) %>% 
  kable_paper("hover")
  

## estimated expat population in wa by country of origin (waves 1-5 only)
fb %>%
  filter(state == "Washington" & wave <= 5) %>%  # looks ok(?) for can, cn, in, mx, ph, vn in waves 6, 7. there's a sizable increase in numbers between waves 5 and 6
  # calculate number of expats by country of origin, by wave
  group_by(wave, date, origin) %>%
  summarize(expat_pop = sum(expat_population), .groups='drop') %>%
  mutate(wave = paste0("wave", wave)) %>%
  pivot_wider(names_from=c(wave, date), values_from=expat_pop) %>%
  kable(format.args=list(big.mark=','), 
        caption = "estimated number of migrants in WA by country of origin, based on FB advertising data") %>%
  kable_paper("hover")

```


## Maps

Maps are a great way to visualize migration patterns. Static maps are the most straight-forward way to showcase a map. 

**ggplot2**

Here, we look at the data for Wave 4 (this is when Hurricane Maria took place). The maps visually show what we found earlier: that the [estimated population size](#expatpopsize) for migrants from Mexico is much larger than those from other countries. To attenuate the numerical difference, the logged population size is displayed on the map. It's also important to remember that the Facebook data only covers 28 countries: it is not the case that there are no migrants from the entire African continent or Latin America.  

```{r}
## plot the data on expat groups
# this is from ggplot2
world_map <- map_data("world")  
#head(world_map)  # this is lat/long point data

## map of fb data for october 2017
fb.pop.counts.w4 <- fb %>% 
  filter(wave == 4) %>%  
  group_by(origin, wave) %>%
  summarize(pop = sum(expat_population), .groups='drop') %>%
  arrange(pop) 

fb.pop.counts.w4 %>%
  # need to keep all country outline data
  right_join(world_map,
            by=c("origin"="region") ) %>%
  # note that mexico has a large number while other countries are much smaller in comparison
  # might need to log scale, or set manual breaks for comparison
  ggplot() +
  geom_polygon(aes(x=long, y=lat, group=group, fill=log(pop)), color="black") +
  coord_fixed(1.3) +
  scale_fill_distiller(na.value="white", direction=1, palette="Reds", name="estimated expats \n in log(population)") +
  labs(subtitle="Country of origin of US migrants, estimated from FB data (Oct 2017)") +
  theme_void() +
  theme(legend.position="bottom")

```


Unlike static maps, interactive maps allow the user to zoom in and out, identify specific features, or query underlying data or indicator (i.e. socioeconomic status). Here, we created a quick and simple interactive map for exploratory data analysis, first using the `mapview` library, then the `tmap` library.


**mapview**

You can find more detailed instructions on how to create an interactive map with [this mapview tutorial](https://bookdown.org/nicohahn/making_maps_with_r5/docs/mapview.html) and the [CRAN documentation](https://cran-r-project.org/web/packages/mapview/mapview.pdf). 

```{r}
## download world outline data (sf)
#require(rnaturalearth)
world <- ne_countries(scale="medium", returnclass="sf")

## create interactive map displaying fb estimates by country of origin
world %>% 
  sf::st_transform(crs=4326) %>%  
  # merge in fb population estimates to shapefile
  left_join(fb.pop.counts.w4, by=c("admin"="origin")) %>%
  select(admin, name, pop_est, continent, region_un, subregion, geometry, wave, pop) %>%
  mapview(zcol="pop",  # which column to map
          map.types=c("OpenStreetMap", "CartoDB.Positron", "Esri.WorldImagery"),  # define which base map(s) to display
          at=c(0, 5e4, 1e5, 2e5, 5e5, 1e6, 2e6, max(fb.pop.counts.w4$pop) ),  # manually define categorical breaks
          col.regions=brewer.pal(7, "YlGn")   # specify color palette              
          )

```

Hovering over one of the shaded countries will give the estimated migrant population size from that country.

**tmap**

The `tmap` library can be used to create both static and interactive maps. Static maps are the default.

```{r, echo=TRUE}
#require(tmap)
## create static tmap

tmap_options(check.and.fix = TRUE)

world %>% 
  sf::st_transform(crs=4326) %>%  
  left_join(fb.pop.counts.w4, by=c("admin"="origin")) %>% 
  tm_shape() +
  tm_fill(col="pop", # which column to plot
          id="name", # what to display
          breaks=c(0, 5e4, 1e5, 2e5, 5e5, 1e6, 2e6, max(fb.pop.counts.w4$pop)),  # manually define breaks
          palette="YlGn") + # specify color palette
  tm_borders()  # add in borders

```

To create an interactive map, specify the mode to `view` using the command `tmap_mode("view")`. To toggle it back to static maps, use `tmap_mode("plot")`. The same code is used to generate the static and interactive maps.

```{r, echo=TRUE}
## create interactive map displaying fb estimates by country of origin
tmap_mode("view")   # interactive mode
#tmap_mode("plot")  # static map (this is the default)

world %>% 
  sf::st_transform(crs=4326) %>%  
  left_join(fb.pop.counts.w4, by=c("admin"="origin")) %>% 
  tm_shape() +
  tm_fill(col="pop", # which column to plot
          id="name", # what to display
          breaks=c(0, 5e4, 1e5, 2e5, 5e5, 1e6, 2e6, max(fb.pop.counts.w4$pop)),  # manually define breaks
          palette="YlGn") + # specify color palette
  tm_borders()  # add in borders


```

Clicking on a country will generate a popup that shows the estimated size of the migrant population from that country.

---

# **American Community Survey (ACS) data**

The code provided by Alexander et al on the GitHub repo indicates they downloaded the 2017 1-year ACS data with sex, age, birthplace (bpl), and state (statefip) variables from [IPUMS](https://usa.ipums.org/usa/). They then perform some data cleaning (the code needed for cleaning the data is available on their Github repo).

We read in the pre-saved ACS 2017 1 year data here: 

```{r}
# read in pre-saved ACS 2017 1-year data
#acs_prop_age <- read_csv("./data/acs_prop_age.csv")  # if cloned/downloaded the github repo, otherwise
acs_prop_age <- read_csv("https://raw.githubusercontent.com/MJAlexander/fb-migration-hurricane-maria/master/data/acs_prop_age.csv")

```


---

# Comparing FB and ACS data

Alexander et al note that Facebook data are not representative of the broader population and it is difficult for the findings to be generalizable. However, the ACS is designed to be nationally representative. Therefore, the comparison of the FB estimates with the ACS estimates enables us to assess this potential bias. 

## National level differences

It is important to recall that the FB data encapsulates only 28 different countries. When looking at this data, we see that there are differences in the number of estimated migrants when compared to that of the ACS. There are 10 countries where the FB estimate is greater than the ACS estimate. For the remaining 18 countries, the ACS estimate is larger than the FB estimate. This indicates a misalignment between the data sources. 

```{r}
## national level estimates of number of "expats" in the acs, compared to fb estimates
acs.fb.comp <- acs_prop_age %>%
  # calculate number of estimated migrants by country of birth  
  group_by(origin, bpl) %>%
  summarize(acs.est = sum(no_mig), .groups='drop') %>%
  select(-bpl) %>%
  # keep only the countries available in the fb data
  filter(origin %in% fb.origins) %>%
  # merge in fb expat data estimates
  right_join(fb.expat %>% select(1:5) %>%  # waves 1-4 "technically" more comparable since covers 2017
               rowwise() %>% mutate(fb.avg = mean(c_across(cols=2:5))) %>% ungroup() , # compute mean estimate over the 4 waves
             by="origin") %>%
  # compute difference between acs and fb estimates
  mutate(diff = acs.est - fb.avg) %>%
  arrange(diff) %>%
  relocate(fb.avg, .after=acs.est) %>%
  relocate(diff, .after=fb.avg)

kable(acs.fb.comp %>% select(1:4), format.args=list(big.mark=',')) %>%
  kable_paper("hover")

```

Note: in the table above, positive values indicate a larger ACS estimate and negative values suggest larger Facebook values.


```{r}
## map the results, to see if potential regional differences
world %>% left_join(acs.fb.comp, by=c("admin"="origin")) %>%
  select(admin, name, pop_est, continent, region_un, subregion, geometry, acs.est, fb.avg, diff) %>%
  mapview::mapview(zcol="diff",  # which column to map
                   map.types=c("OpenStreetMap", "CartoDB.Positron", "Esri.WorldImagery"),  # define which base map(s) to display
                   at=c(min(acs.fb.comp$diff), -1e4, 0, 1e5, 5e5, 1e6, max(acs.fb.comp$diff)),  # manually define categorical breaks
                   col.regions=brewer.pal(6, "RdBu")
                   )

```


## State level differences

When examining differences in the PR expat population by state, the authors exclude North Carolina from the analysis. They explain that it is in an anomaly. They also examined only states with 18,000+ PRs to avoid rounding issues. For these 11 states, while the ACS and Facebook data do yield different migrant population estimates at the state level, most of them are comparable. However, the disparity between the two data sets are slightly greater (indicating a greater misalignment) for Connecticut and New Jersey. The states with the largest PR populations are Florida, New York, and Pennsylvania.  

```{r}
# 4. Top states based on PR population --------------------------------

# table of expat populations and proportions in facebook and acs
# remove North Carolina because there seems to be an anomaly in the data 
# filter to include only states above 18000 to avoid rounding issues. 

pr.state <- fb %>% 
  # authors do not include north carolina in their analysis for some reason (??)
  filter(!state %in% c("North Carolina")) %>% 
  # keep fb wave 4 data only, and origin country = pr
  filter(wave==4, origin=="Puerto Rico") %>% 
  # merge in acs data on pr population
  left_join(acs_prop_age %>% 
              filter(origin == "Puerto Rico"),
            by=c("state", "origin", "age_group", "sex")) %>%   
  # calculate for each state:
  group_by(state) %>% 
  summarise(expat_population = sum(expat_population),  #  total fb pr expat population
            total_pop = sum(facebook_population),   # total fb population 
            expat_population_acs = sum(no_mig),  # total acs pr migrant pop
            expat_proportion_acs = round(sum(prop_pop), 3)  # proportion of migrants in state
            ) %>% 
  # calculate fb expat proportion (out of all fb users)
  mutate(expat_proportion = round(expat_population/total_pop, 3)) %>% 
  select(state, expat_population, expat_population_acs, expat_proportion, expat_proportion_acs) %>% 
  # arrange in decreasing numbers of fb pr migrants by state
  arrange(-expat_population) 

# states where pr expat pop > 18000 (12, if north carolina not dropped)
pr.state %>% 
  filter(expat_population>18000) %>%
  kable(format.args=list(big.mark=',')) %>%
  kable_paper("hover")

```

## Age distribution
(Figure 1 from the paper)

When examining the age distribution of male migrants in 9 states, it is evident that the ACS has an older age distribution while the Facebook distributions are greater for the under-30 age groups. The ACS trends differ in all age groups.  

```{r}
# 7. age change -----------------------------------------------------------

## compare age distributions (figure 1 in the paper)
acs_prop_age %>% 
  # pr individuals in the acs only
  filter(origin=="Puerto Rico") %>% 
  # merge in fb data on pr expats from wave 1 (jan 2017)
  left_join(fb %>% filter(wave==1, origin=="Puerto Rico")) %>% 
  # calculate proportion of expats in each age group out of total fb pop per sex-state combination
  group_by(state, sex) %>% 
  mutate(expat_prop = expat_population/sum(expat_population, na.rm = T)) %>% 
  ungroup() %>%
  select(state, sex, age_group, prop_mig, expat_prop) %>% 
  # look at 9 selected states, males
  filter(sex==1, state %in% c("Florida", "New York", "New Jersey", 
                              "Pennsylvania", "Connecticut", "Illinois",
                              "California", "Texas", "Massachusetts")) %>% 
  ggplot(aes(age_group, prop_mig)) + 
  geom_line(lwd = 0.8) + 
  geom_line(aes(age_group, expat_prop), color = "red", lty = 2, lwd = 0.8) + 
  facet_wrap(~state)+
  theme_bw(base_size = 14) + 
  ylab("proportion") + xlab("age group")

```

(FB data represented by the red dashed line; ACS data by the solid black line.)

## Sex ratio

The following Table is Table 1 in the paper. It compares the sex ratio estimated from the 2017 ACS with the FB data from wave 1 (Jan 2017). The Facebook sex ratios are generally larger than the ACS sex ratios, indicating higher proportions of men than women. The visualization compares the difference in estimated sex ratios by data source for selected states. 

```{r}
# 8. sex ratios -----------------------------------------------------------

#https://stats.stackexchange.com/questions/21167/standard-error-of-a-ratio

## sex ratios in ACS vs fb, by state  
sr.nat <- acs_prop_age %>% 
  mutate(pr = ifelse(origin== "Puerto Rico", 1, 0)) %>% 
  group_by(state, pr, sex) %>%
  summarise( no_mig = sum( no_mig, na.rm = T)) %>% 
  group_by(state, pr) %>% 
  # acs sex ratio (2017)
  summarise(sr_acs = no_mig[sex==1]/no_mig[sex==2]) %>% 
  group_by(state) %>% 
  filter(state %in% c("Florida", "New York", "New Jersey", 
                      "Pennsylvania", "Illinois", "California", "Connecticut",
                      "Texas", "Massachusetts", "Ohio", "Georgia")) %>% 
  left_join(fb %>% 
              mutate(pr = ifelse(origin== "Puerto Rico", 1, 0)) %>% 
              mutate(expat_population = ifelse(state=="Connecticut"&age_group==35&wave==4&sex==1&pr==1, 4300, expat_population)) %>% 
              group_by(wave, state, pr, sex) %>%
              summarise(expat_population = sum(expat_population, na.rm = T)) %>% 
              group_by(wave, state, pr) %>% 
              summarise(sr_fb = expat_population[sex==1]/expat_population[sex==2]) %>%  # fb sex ratio (m:f)
              filter(state %in% c("Florida", "New York", "New Jersey", 
                                  "Pennsylvania", "Illinois", "California", "Connecticut",
                                  "Texas", "Massachusetts", "Ohio", "Georgia"), wave<6)) %>%  ungroup() %>%
  filter(pr==1, wave==1) %>% select(-pr, -wave)  # oct 2017

## table comparing sex ratio in acs data vs fb data (table 1 in the paper)
sr.nat %>%
  kable(digits=3, col.names=c("state", "ACS sex ratio", "Facebook sex ratio")) %>%
  kable_paper()

## plot difference in sex ratio by data source, for select states
sr.nat %>%
  pivot_longer(2:3, names_to="source", values_to="sr") %>%
  separate(source, into=c(NA, "source")) %>%
  arrange(source, sr) %>%
  ggplot(aes(y=fct_reorder2(state, source, sr), # sort values
             x=sr, color=source)) +
  geom_point() +
  geom_vline(xintercept=1, linetype=2) +
  labs(x="sex ratio", y="state") +
  scale_y_discrete(limits=rev) +  # for reversing the order of the states in the plot
  theme_bw() +
  theme(legend.position="bottom")

```

---

# **Analysis: Difference-in-Differences estimation**

Difference-in-differences models are panel data methods applied when we are attempting to understand the effect of a particular treatment. Difference-in-differences models are used when some groups are exposed to a treatment and others are not. In this study, Alexander et al used difference-in-differences to estimate the percent change in Puerto Rican migrants in the continental US after Hurricane Maria. They choose this method to overcome issues of non-representativness and fluctuations in the data over time (indpendent of migration events). 

Difference-in-difference models haven been be used to study the effects of a [particular policy](https://diff.healthpolicydatascience.org/), in papers focused on the [economics of education](https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/B978012815391800001X), and [labor economics](https://www-sciencedirect-com.offcampus.lib.washington.edu/topics/economics-econometrics-and-finance/difference-in-differences). 

In this paper, the diff-in-diff estimator is
$$
\pi_{g} = \frac{P_{g}^{PR}(5) - P_{g}^{PR}(4)}{P_{g}^{PR}(4)} - \frac{P_{g}^{c}(5) - P_{g}^{c}(4)}{P_{g}^{c}(4)} 
$$
where $g$ is the group of interest (age, sex, country of origin, state of residence). The authors are interested in assessing the difference in the size of the migrant population from Puerto Rico between waves 4 and 5, relative to the change in the size of the migrant population from all other countries, sans Mexico.

The diff-in-diff approach assumes the control group, which is not subject to the treatment (here, Hurricane Maria), will continue to exhibit similar responses, relative to the treatment group. The difference between groups is the effect of the treatment. The treatment group is Puerto Rican migrants, while the control group is composed of migrants from all other countries aside from Mexico, which is assumed to follow different migration motivations. Effectively, in using this estimator, we are comparing the change in the size of the PR migrant population in the US between waves 4 and 5 with the change in population for all other migrants (but not Mexican immigrants), also between waves 4 and 5.


## National diff-in-diff estimate of out-migration from Puerto Rico

Using the difference-in-difference model, Alexander et al find a 17% increase in the number of Puerto Rican migrants present in the continental US between October 2017 and January 2018. 

```{r}
exp_prop_usa <- fb %>% 
  # exclude migrants from puerto rico or mexico
  filter(origin != "Puerto Rico", origin != "Mexico") %>%
  # calculate number of expats in each wave
  group_by(wave) %>% 
  summarise(expat = sum(expat_population)) %>% 
  # add in total number of fb users per wave
  left_join(total_fb_pop, by="wave") %>% 
  # calculate proportion of expats out of total fb population
  mutate(prop = expat/fb) %>% 
  # merge this information back into fb data
  left_join(fb %>% 
              # calculate number of pr expats in the us, per wave
              filter(origin=="Puerto Rico") %>% 
              group_by(wave) %>% 
              summarise(expat_pr = sum(expat_population)),
            by="wave") %>% 
  # calculate proportions
  mutate(prop_pr = expat_pr/fb,  # proportion of pr expat users among total fb users
         var_prop = prop*(1-prop)/(fb/1000),  # binomial variance for all other migrants (excludes mx) 
         var_prop_pr = prop_pr*(1-prop_pr)/(fb/1000)) %>%  # binomial variance for pr migrants
  # calculate diff in diff  
  mutate(prop_diff = (prop - lag(prop))/lag(prop),  # calculate difference in proportion of all other migrants between waves
         se_diff = sqrt(var_prop + lag(var_prop)),  
         prop_pr_diff = (prop_pr - lag(prop_pr))/lag(prop_pr),  # calculate difference in pr migrants
         se_diff_pr = sqrt(var_prop_pr + lag(var_prop_pr)),  
         diff_in_diff = (prop_pr_diff - prop_diff)*100,  # calculate diff-in-diff estimator, x100 to get pct
         se_diff_diff = sqrt(se_diff^2 + se_diff_pr^2)*100)  # calculate se for diff-in-diff estimator

# pull out diff-in-diff estimator (+se, 95% ci) between waves 4 and 5
res <- exp_prop_usa %>% filter(wave==5) %>% 
  select(percent_increase = diff_in_diff, se = se_diff_diff) %>% 
  # calculate ~95% ci
  mutate(ci_lower = percent_increase - (2*se), 
         ci_upper = percent_increase + (2*se))

res %>%
  kable(digits=2) %>%
  kable_paper()

```

## National estimate of PR out-migrants to the US

The authors then take the estimate percent changes obtained from the difference-in-differences model and multiplies them by the relevant populations reported in the ACS. This enables us to estimate the change in the *number* of migrants.

$$\hat{M}_{g} = \hat{\pi}_{g} \times P_{g}^{ACS}$$
where $\hat{M_{g}}$ is the estimated number of migrants in group $g$, $\hat{\pi}_{g}$ is the proportional change in PR migrants in group $g$, and $P_{g}^{ACS}$ is the estimated number of PR migrants in group $g$ in the ACS. 

(The 2017 ACS indicates there were 1,087,557 Puerto Ricans residing in the US.)


```{r}
acs_prop_age %>% 
  filter(origin=="Puerto Rico") %>% 
  # calculate mean number of migrants, using previously calculated percentage of movers and 95% bounds
  summarise(mean = sum(no_mig)*res$percent_increase/100, 
            lower = sum(no_mig)*res$ci_lower/100, 
            upper = sum(no_mig)*res$ci_upper/100) %>%
  kable(format.args=list(big.mark=",")) %>%
  kable_paper()

```



## State diff-in-diff estimates of in-migrants from PR

The authors then repeat the same approach at the state level, presenting results for states where the estimated PR population in the FB data exceeds 18,000.

Data note: In the code, `expat_pop` = 20  seems to be some kind of bottom-code or placeholder for missing data or small counts (20 is the minimum count in the data). The authors only examine counts greater than 20 for their analysis, but this is not explained in the paper or the code. This may be due to limited documentation with the Facebook data. Additionally, North Carolina is also excluded from the analysis (there isn't an explanation as to why).

```{r}
# 5a. State by state analysis (table) ----------------------------------------------

## calculate approximate population per state, by state and wave, using fb data
total_fb_pop_state <- fb %>% filter(origin=="Mexico") %>% 
  group_by(wave, state) %>% 
  summarise(fb = sum(facebook_population), .groups='drop')

## calculate expat population proportions by waves
ex_props <- fb %>%
  filter(!state %in% c("North Carolina")) %>%  # exclude north carolina (?)
  # exclude all expats from puerto rico, mexico, where expat_pop = 20
  filter(origin!="Puerto Rico", origin!= "Mexico", expat_population>20) %>%
  # calculate for each state and fb wave, total number of expats from all origins besides mx, pr
  group_by(state, wave) %>%
  summarise(expat = sum(expat_population), .groups='drop') %>%
  # merge back fb population by state
  left_join(total_fb_pop_state,
            by=c("state", "wave")) %>% 
  # calculate proportion of fb expats in each state
  mutate(prop = expat/fb) %>% 
  group_by(state) %>%
  # merge in data on pr migrants from the fb data
  left_join(fb %>%
              filter(origin=="Puerto Rico") %>%
              # calculate number of pr expats in each state, by wave
              group_by(state, wave) %>%
              summarise(expat_pr = sum(expat_population), .groups='drop'),
            by=c("state", "wave") ) %>% 
  # calculate for each state
  mutate(prop_pr = expat_pr / fb,  # proportion of pr expats out of total fb users
         var_prop = prop*(1-prop)/(fb/1000),  # variance for proportion of expats out of total fb users 
         var_prop_pr = prop_pr*(1-prop_pr)/(fb/1000)) %>%  # variance for proportion of pr users out of total fb users
  # calculate diff in diff estimator for each state
  # this will throw some warning messages for several states (ak, nd, sd, vt, wy)
  # (from the huge difference in expat_pop between waves 5 and 6, 6 and 7)
  group_by(state) %>%
  mutate(prop_diff = (prop - lag(prop))/lag(prop),  # difference in proportion of expats between waves
         se_diff = sqrt(var_prop + lag(var_prop)),  
         prop_pr_diff = (prop_pr - lag(prop_pr))/lag(prop_pr),  # difference in pr expats between waves
         se_diff_pr = sqrt(var_prop_pr + lag(var_prop_pr)),  
         diff_in_diff = (prop_pr_diff - prop_diff)*100,  # diff-in-diff estimator * 100 (to get percent)
         se_diff_diff = sqrt(se_diff^2 + se_diff_pr^2)*100) %>%  
  ungroup()

# table with percent increases and numbers
ex_props %>%  
  filter(expat_pr>18000, !is.na(diff_in_diff), wave==5) %>% 
  select(state, prop_pr, diff_in_diff, se_diff_diff) %>% 
  mutate(percent_increase = round(diff_in_diff, 1),  
         percent_lower = percent_increase - (2*se_diff_diff), 
         percent_upper = percent_increase + (2*se_diff_diff)) %>%   
  select(state, percent_increase, percent_lower, percent_upper) %>% 
  # arrange states by large -> small pct increase in pr expats  
  arrange(-percent_increase) %>% 
  # add in acs data
  left_join(acs_prop_age %>% 
              filter(origin=="Puerto Rico") %>% 
              # calculate nubmer of pr migrants by state
              group_by(state) %>%  
              summarise(mig = sum(no_mig))) %>% 
  # calculated estimated number of pr migrants in the us using fb pct estimate and acs pop est
  mutate(estimated_number = round(mig*percent_increase/100),
         number_lower = round(mig*percent_lower/100),
         number_upper = round(mig*percent_upper/100)) %>% 
  select(-mig) %>% 
  arrange(-estimated_number) %>%
  kable(format.args=list(big.mark=','), digits=2) %>%
  kable_paper("hover")


```

The estimates suggest post Hurricane Maria, many Puerto Ricans were moving to the states that have an existing population of Puerto Rican migrants. This can indicate community based migration/chain migration. 


### Map
(Figure 2 in the paper)

The largest increase of Puerto Rican migrants is in Florida. 

```{r}
# 5b. State by state analysis (map) ---------------------------------------

to_map <- ex_props %>%  
  # keep states/waves with over 18k pr fb expats
  filter(expat_pr>18000) %>% 
  # keep results from wave 5 (estimates difference between waves 4, 5)
  filter(!is.na(diff_in_diff), wave==5) %>% 
  select(state, prop_pr, diff_in_diff) %>% 
  mutate(percent_increase = round(diff_in_diff, 1)) %>% 
  select(state, percent_increase) %>% 
  arrange(-percent_increase) 

## load us map outline (part of ggplot2)
states_map <- map_data("state")

# create map for illustrating percent increase by state (choropleth map - this is figure 2 in the paper)
states_map %>% 
  # merge in data on percent increase by state
  left_join(to_map %>% 
              # format for merging with map data
              mutate(region = str_to_lower(state)) ) %>% 
  ggplot() + 
  geom_polygon(aes(x = long, y = lat, fill = percent_increase, group = group), color = "black") + 
  coord_fixed(1.3) +  # define aspect ratio
  theme_void() +  # removes lat/long axes
  scale_fill_distiller(na.value = "white", direction = 1, palette = "Reds", name = "percent increase")

```


## Age-specific estimates of out-migrants from PR to the US
(Figure 3 in the paper)

In terms of who migrated out of Puerto Rico after Hurricane Maria, there is an increase in those in the 15-29 year age range. There were fewer people above the age of 35 migrating out of Puerto Rico in a post-disaster context. 

```{r}
# pr pop vs all other migrant pop, according to fb data
# (non-pr includes mexico)
d_tot <- 
  fb %>% 
  mutate(pr = ifelse(origin== "Puerto Rico", 1, 0)) %>% 
  # change pr expat pop of those age 35-39, male in connecticut to 4300. as given in data: 20. in wave 3 the number is 4300
  mutate(expat_population = ifelse(state=="Connecticut"&age_group==35&wave==4&sex==1&pr==1, 4300, expat_population)) %>% 
  # calculate total pr/non-pr migrant population for each age group, by wave
  group_by(pr, age_group, wave, date) %>% 
  summarise(expat_population = sum(expat_population), .groups='drop') %>%  # 7 waves x 9 age group x 2 migrant groups = 126 unique combinations
  group_by(wave, pr) %>% 
  # calculate proportion of age group out of total fb expat (pr, non-pr) population in that wave (separate for pr, non-pr)
  mutate(exp_prop = expat_population/sum(expat_population),  
         var_prop = exp_prop*(1-exp_prop)/expat_population*10) %>%  # associated variance 
    ungroup()  

```


```{r, include=F, echo=F}
# diff-in-diff by each age group, national (table)
d_tot %>% 
  group_by(age_group) %>%
  # calcuating diff-in-diff estimator
  summarise(did = ((exp_prop[wave==5&pr==1]-exp_prop[wave==4&pr==1])*100 - # pr pop
                     (exp_prop[wave==5&pr==0]-exp_prop[wave==4&pr==0])*100),  # non-pr pop
            se_did = sqrt(sqrt(var_prop[wave==5&pr==1]+var_prop[wave==4&pr==1])^2 + sqrt(var_prop[wave==5&pr==0]+var_prop[wave==4&pr==0])^2),  # associated se
            lower = did - (2*se_did), upper = did + (2*se_did))  # approx 95% ci

```


```{r}
## plot results (figure 3 in the paper, but with a confidence interval)
# line plot
d_tot %>% 
  group_by(age_group) %>%
  summarise(did = ((exp_prop[wave==5&pr==1]-exp_prop[wave==4&pr==1])*100 - (exp_prop[wave==5&pr==0]-exp_prop[wave==4&pr==0])*100),
            se_did = sqrt(sqrt(var_prop[wave==5&pr==1]+var_prop[wave==4&pr==1])^2 + sqrt(var_prop[wave==5&pr==0]+var_prop[wave==4&pr==0])^2),
            lower = did - (2*se_did), upper = did + (2*se_did) ) %>% 
  ggplot(aes(age_group, did)) + geom_line() + geom_hline(yintercept = 0) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) + 
  xlab("Age group") + ylab("Change (percentage points)") + 
  theme_bw(base_size = 14)

```


```{r, include=F, echo=F}
# barplot
d_tot %>% 
  group_by(age_group) %>%
  summarise(did = ((exp_prop[wave==5&pr==1]-exp_prop[wave==4&pr==1])*100 - (exp_prop[wave==5&pr==0]-exp_prop[wave==4&pr==0])*100),
            se_did = sqrt(sqrt(var_prop[wave==5&pr==1]+var_prop[wave==4&pr==1])^2 + sqrt(var_prop[wave==5&pr==0]+var_prop[wave==4&pr==0])^2),
            lower = did - (2*se_did), upper = did + (2*se_did)) %>% 
  ggplot(aes(age_group, did)) + 
  geom_bar(stat = "identity", fill = "#772D8B") + geom_hline(yintercept = 0) + 
  #geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) + 
  xlab("Age group") + ylab("") + 
  #labs(title = "Change in percentage shares of all Puerto Rican migrants in the US", subtitle = "from October 2017 to January 2018")+
  theme_bw(base_size = 14)

```

## Change in sex ratio of out-migrants from PR, by state
(Table 3 in the paper)

Pre- and post-hurricane, there were larger numbers of men than women in all nine states. However, there were some states (i.e. Texas) that show an increase of male migrants, while other states (i.e. Pennsylvania) saw an increase of female migrants. 

```{r}
## change in sex ratio between waves in the fb data, by state (table 3)
fb %>% 
  mutate(pr = ifelse(origin== "Puerto Rico", 1, 0)) %>% 
  mutate(expat_population = ifelse(state=="Connecticut"&age_group==35&wave==4&sex==1&pr==1, 4300, expat_population)) %>% 
  left_join(total_fb_pop_state) %>% 
  group_by(wave, state, pr, sex) %>%
  summarise(expat_population = sum(expat_population, na.rm = T),  # total expat_pop by combination
            fb = fb[row_number()==1],  # keep the first value in the fb column for each wave/sex combination
            prop = expat_population/fb,  # calculate pr expat pop as proportion of total fb pop
            var_prop = prop*(1-prop)/(fb/1000)) %>% # calculate variance for the proportion
  group_by(wave, state, pr) %>% 
  summarise(sr = expat_population[sex==1]/expat_population[sex==2],   # calculate sex ratio in fb data
            se = (prop[sex==1]/prop[sex==2])^2*(var_prop[sex==1]/(prop[sex==1]*100)^2 + var_prop[sex==2]/(prop[sex==2]*100)^2)) %>%  # calculate associated se (see stacks link from sex ratio section)
  group_by(state) %>% 
  # calculate difference in sex ratio between waves 4/5 and between pr, non-pr
  mutate(sr_diff = (sr[wave==5&pr==1]-sr[wave==4&pr==1])/sr[wave==4&pr==1] - (sr[wave==5&pr==0]-sr[wave==4&pr==0])/sr[wave==4&pr==0], 
         # calculate se for difference in sex ratio
         se_diff = sqrt(sqrt(se[wave==5&pr==1]^2+se[wave==4&pr==1]^2) + sqrt(se[wave==5&pr==0]^2+se[wave==4&pr==0]^2))) %>%   
  filter(state %in% c("Florida", "New York", "New Jersey", 
                      "Pennsylvania", "Illinois", "Ohio", "Connecticut",
                      "Texas", "Massachusetts", "Georgia", "California"), wave == 4, pr==1) %>%
  mutate(sex_ratio=round(sr,3),
         change = round(sr_diff,3), 
         lower = round(sr_diff - (2*se_diff),3), 
         upper = round(sr_diff + (2*se_diff),3)) %>% 
  select(state, starts_with("sex_ratio"), change, lower, upper) %>% 
  arrange(-change) %>%
  kable() %>%
  kable_paper()

```

---

# **Return migration**

## National diff-in-diff estimate of out-migration from the US to Puerto Rico

In estimating return migration, Alexander et al specifically look at the decrease in the Puerto Rican population in the continental US after the hurricane. However, it doesnt seem like there is enough information to suggest that an inference about where the migrants who originate from Puerto Rico move to after their time in the US is appropriate. While it is likely that it is possible that they return, this migration pattern is inferred.

Using this assumption the results indicate a 1.8 percent of Puerto Ricans that return to Puerto Rico after the hurricane.

Data note: Authors recognize that there is a rounding issue after Wave 5, thus keeping only those records with expat_population over 1000)

```{r}
# 6. Return migration -----------------------------------------------------

# national
ex_props_post_nat <- fb %>%
  filter(origin!="Puerto Rico", origin!= "Mexico", expat_population>1000) %>%  # rounding issue after wave 5
  group_by(wave) %>%
  summarise(expat = sum(expat_population)) %>%  # calculate number of expats per wave
  # merge in total fb population in the us
  left_join(total_fb_pop, by="wave") %>% 
  # calculate proportion of expats in the total fb population
  mutate(prop = expat/fb) %>% 
  left_join(fb %>%
              filter(origin=="Puerto Rico", expat_population>1000) %>%
              # caclulate fb pr expat population by wave
              group_by(wave) %>%
              summarise(expat_pr = sum(expat_population)),
            by="wave") %>% 
  mutate(prop_pr = expat_pr / fb,  # proportion of pr expats out of total fb population
         var_prop = prop*(1-prop)/(fb/1000),  # variance of expats out of total fb population
         var_prop_pr = prop_pr*(1-prop_pr)/(fb/1000)) %>%  # variance of pr expats out of total fb population
  mutate(prop_diff = (prop - lag(prop))/lag(prop),  # calculate difference between waves (all expats, -mx)
         se_diff = sqrt(var_prop + lag(var_prop)),  # se for the difference between waves
         prop_pr_diff = (prop_pr - lag(prop_pr))/lag(prop_pr),  # calculate difference between waves, pr only
         se_diff_pr = sqrt(var_prop_pr + lag(var_prop_pr)),  # se for the difference between waves
         diff_in_diff = (prop_pr_diff - prop_diff)*100,  # calculate diff-in-diff estimator (in percent)
         se_diff_diff = sqrt(se_diff^2 + se_diff_pr^2)*100)  # se for diff-in-diff estimator (in percent)

res_return <- ex_props_post_nat %>% 
  # examine difference between waves 5 and 6
  filter(wave==6) %>% 
  select(diff_in_diff, se_diff_diff) %>% 
  rename(percent_increase = diff_in_diff, se = se_diff_diff) %>% 
  mutate(ci_lower = percent_increase - (2*se), ci_upper = percent_increase + (2*se))

res_return %>%
  kable(digits=2, caption="estimated percentage change") %>%
  kable_paper()

## use these percentages with acs data to estimate number of return migrants to pr (at national level)
acs_prop_age %>% filter(origin=="Puerto Rico") %>% 
  ungroup() %>%  
  summarise(mean = sum(no_mig)*res_return$percent_increase/100, 
            lower = sum(no_mig)*res_return$ci_lower/100, 
            upper = sum(no_mig)*res_return$ci_upper/100) %>%
  kable(format.args=list(big.mark=','), caption="estimated number of return migrants") %>%
  kable_paper()

```

## State diff-in-diff estimates of out-migrants from the US to PR

Making the same assumption as above, Alexander et al also find that there is a decrease in the Puerto Rican population in Florida and Texas (the states with the greatest in-migration during the hurricane). Other states like California and New York saw an increase of Puerto Rican migrants during this time.

```{r}
# by state
ex_props_post <- fb %>%
  # all other expats, but pr and mx
  filter(origin!="Puerto Rico", origin!= "Mexico", expat_population>1000) %>%  # rounding issue after wave 5
  group_by(state, wave) %>%
  summarise(expat = sum(expat_population), .groups='drop') %>%  # calculate number of expats in each state per wave
  # merge back in total fb population
  left_join(total_fb_pop_state) %>% 
  mutate(prop = expat/fb) %>%  # calculate expat proportion out of total fb population
  group_by(state) %>%
  left_join(fb %>%
              filter(origin=="Puerto Rico") %>%
              group_by(state, wave) %>%
              summarise(expat_pr = sum(expat_population))) %>%  # number of pr expats per state
  # calculate proportions
  mutate(prop_pr = expat_pr / fb,  # pr expats as a proportion of total fb population
         var_prop = prop*(1-prop)/(fb/1000),  # variance for all expats as proportion of total fb population
         var_prop_pr = prop_pr*(1-prop_pr)/(fb/1000)) %>%  # variance for pr expats as proportion of total fb pop
  # calculate diff in diff for each state and wave
  arrange(state, wave) %>% 
  group_by(state) %>%
  mutate(prop_diff = (prop - lag(prop))/lag(prop),  # difference between waves (all expats -mx, -pr)
         se_diff = sqrt(var_prop + lag(var_prop)),  # se for difference in "all" expats between waves
         prop_pr_diff = (prop_pr - lag(prop_pr))/lag(prop_pr),  # difference in pr expats between waves
         se_diff_pr = sqrt(var_prop_pr + lag(var_prop_pr)),  # se for difference in pr expats between waves
         diff_in_diff = (prop_pr_diff - prop_diff)*100,  # diff-in-diff estimator (in percent)
         se_diff_diff = sqrt(se_diff^2 + se_diff_pr^2)*100) %>%  # se for diff-in-diff estimator
  ungroup()
# notice that only 33 states

ex_props_post %>% 
  # keep only state-wave combinations where total pr expats greater than 18k (leaves 14 states [if nc included])
  #  and diff-in-diff estimators from wave 6 (this is the difference b/w waves 5 and 6)
  filter(expat_pr>18000, !is.na(diff_in_diff), wave == 6)  %>% 
  select(state, prop_pr, diff_in_diff, se_diff_diff) %>% 
  mutate(percent_change = round(diff_in_diff, 1),
         percent_lower = percent_change - (2*se_diff_diff), 
         percent_upper = percent_change + (2*se_diff_diff) ) %>% 
        select(state, percent_change, percent_lower, percent_upper) %>% 
        arrange(percent_change) %>% 
  # merge in acs estimates of pr indivdiuals in the us by state
  left_join(acs_prop_age %>% 
              filter(origin=="Puerto Rico") %>% 
              group_by(state) %>%  
              summarise(mig = sum(no_mig))) %>%  # calculate number of pr individuals in each state
  # calculate estimated number of pr return migrants + associated ci
  mutate(estimated_number = round(mig*percent_change/100),
         number_lower = round(mig*percent_lower/100),
         number_upper = round(mig*percent_upper/100)) %>% 
  # remove acs pr estimate per state
  select(-mig) %>% 
  # arrange states in order by estimated number of pr return migrants
  arrange(estimated_number) %>%
  kable(format.args=list(big.mark=','), digits=3, 
        caption="estimated return migrants to Puerto Rico for select states") %>%
  kable_paper()


```